---
title: "Do Vote-By-Mail A Partisan Matter?: Lessons from linking NC elections data"
class: "Data Analysis for the Social Sciences"
professor: "Gregory Eirich"
author: "Yun Choi"
date: "12/18/2021"
output: 
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: true
    collapsed: false
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, message = FALSE)

library(tidyverse)
library(janitor)
library(stargazer)
library(psych)
library(DT)
library(car)
library(jtools)
```

# 1. Research Topic
Throughout the 2020 U.S. presidential election, then-President Donald Trump continued his attacks on mail-in voting. Although the extent of his influence remains unclear, only 30% of Republicans voted by mail despite nationwide polling place closures and the health risk of casting an in-person ballot during the COVID-19 pandemic. In comparison, nearly 60% of Democrats cast a mail-in ballot. Such a disparity raises the question of whether vote-by-mail (VBM) is indeed a partisan issue and less accepted by Republicans, even when controlling for other factors that influence one’s preference for VBM. Survey studies have found that Republicans are less likely to VBM than Democrats, but they did not consider other factors that might influence their preference of voting method. 

# 2. Hypothesis
In this paper, I examine the following hypothesis: (1) being a registered Republican (independent variable) decreases the probability of VBM (dependent variable), even when controlling for voter age, race, sex, and distance to a designated polling location (controlled variables); and (2) being white has differential impacts on the probability of VBM across the political spectrum. I hypothesize that within the Republican Party, being white intensifies disapproval against VBM. On the other hand, within the Democratic Party, being white increases the probability of VBM. 

I control for voter age and distance to a polling place because age and travel distance are two primary factors in deciding whether to vote-by-mail (VBM). The older the voter, and the longer the travel distance, the higher the voter's incentive to VBM and save the trip. I also control for voter race and gender because white, and male voters are likely to have a higher socioeconomic status than non-white, and female voters, respectively, which is also one of the strong predictors for a high probability of VBM.

My first hypothesis will be falsified if the probability of VBM is constant across the political spectrum, net of all the controlled variables. That means the disparity in the probability of VBM derives from some or all of the variables I control for. My second hypothesis will be falsified if being white influences the probability of VBM in the same direction across the political spectrum. That means being white has the same effect on the probability of VBM, regardless of their political affiliation. 

# 3. Data Description
My ideal data would show each voter's demographic and socioeconomic characteristics, political affiliation, and preference for the voting method. However, few states have made such personal information public due to privacy reasons. I have chosen North Carolina among all states because its voter file has individual-level demographic information such as race, ethnicity, and sex, unavailable in most states. And it is open to the public.

By linking various elections datasets – voter registrations, voter history, and polling places – from the North Carolina State Board of Elections, I have created a dataset that shows the demographic, political, and voting preference information for each voter in North Carolina. I filtered 6.1 million who voted for the 2020 presidential election only, and randomly selected 10 samples from each of nearly 2,600 precincts in North Carolina. That way, all precincts in different parts of the state are represented. That has left me with about 26,000 samples in total, which is about 0.4% of the entire 2020 presidential voters in North Carolina. 

I focus on the 2020 presidential election because during the period, VBM was most needed due to the global pandemic, but also caused a strong conflict between Democrats and Republicans. I selected a subset of voters due to the limits of computing power required for geocoding. It was infeasible to geocode residential and polling location addresses and calculate the distance between the two points for each voter. 

Below is the list of variables I use in this paper:
•	**Dependent:** VBM (binary - '1' if voted by mail; '0' if not) I created a new dummy variable showing whether each voter voted by mail during the 2020 presidential election. Those who voted by mail have a ‘1’ as the value, while the others have a ‘0’.
•	**Independent:** Rep (binary - '1' if registered Republican; '0' if not) I created a new dummy variable showing whether each voter was a registered Republican, based on voter party affiliation status during the election. Registered Republicans have a ‘1’ as the value, while the others have a ‘0’.
•	**Controlled 1:** Age (continuous - age as of 2020) I created a new continuous variable based on the date of birth column in the voter registration data. I calculated the voter age as of the date of the 2020 presidential election. 
•	**Controlled 2:** White (binary - '1' if non-Hispanic white; '0' if not) I created a new dummy variable that shows whether the voter is non-Hispanic, based on two preexisting columns – race and ethnicity. Non-Hispanic whites have a ‘1’ as the value, while the others have a ‘0’.
•	**Controlled 3:** Male (binary - '1' if male; '0' if not) I created a new dummy variable that shows whether the voter is male, based on the preexisting ‘sex’ column. Males have a ‘1’ as the value, while the others have a ‘0’.
•	**Controlled 1:** TravelDistance (continuous – travel distance b/w residence and polling location in miles) I geocoded each voter’s residence and designated polling place and calculated the geodesic distance between two points using the ‘gdist’ function in the ‘Imap’ package in R because of the ease of its calculation. The function uses the Vincenty inverse formula for ellipsoids. However, the geodesic distance would differ from actual travel distance over a road network. I intend to recalculate the travel distance between the two points using the ‘gmapdistance’ function in the identically named R package. The function uses the Google Maps Distance Matrix API to compute the distance between two points in meters.
•	**Interaction:** Rep * White (binary – ‘1’ only if non-Hispanic white Republican)

```{r}
# Read data
final_dist_no_outliers <- read_csv("final_dist_no_outliers.csv")

data_for_models <- final_dist_no_outliers %>% 
  mutate(Rep = if_else(voted_party_cd == "REP", 1, 0), 
         VBM = if_else(voting_method == "ABSENTEE BY MAIL", 1, 0), 
         White = if_else(race_code == "W" & ethnic_code == "NL", 1, 0), 
         Age = 2021 - birth_year, 
         Male = if_else(gender_code == "M", 1, 0),
         TravelDistance = distance) 

all_three_joined_desc <- all_three_joined %>%
    mutate(Rep = if_else(voted_party_cd == "REP", 1, 0), 
         VBM = if_else(voting_method == "ABSENTEE BY MAIL", 1, 0), 
         White = if_else(race_code == "W" & ethnic_code == "NL", 1, 0), 
         Age = 2021 - birth_year, 
         Male = if_else(gender_code == "M", 1, 0)) 
```

* Below is the list of variables I explore in this paper. 
 + Dependent: VBM (binary - '1' if voted by mail; '0' if not) 
 + Independent: Rep (binary - '1' if voted for the Republican party; '0' if not)
 + Controlled 1: TravelDistance (continuous - travel distance b/w residence and polling location in miles)
 + Controlled 2: Age (continuous - age as of 2020)
 + Controlled 3: White (binary - '1' if non-Hispanic white; '0' if not)
 + Controlled 4: Male (binary - '1' if male; '0' if not)
 + Interaction: Rep * White

# 4. Descriptive Statistics

```{r}
all_three_joined_desc %>%
  select(VBM, Rep, Age, White, Male) %>%
  describe() %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  select(-vars) %>%
  kable(caption = "TABLE 1: Population Variables")
```


```{r}
data_for_models %>%
  select(VBM, Rep, Age, White, Male, TravelDistance) %>%
  describe() %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  select(-vars, -trimmed, -mad, -skew, -kurtosis) %>%
  kable(caption = "Sample Variables")
```

TABLE 1 shows the mean, standard deviation, and range for each variable for the population (6.1 million North Carolinians voted for the 2020 presidential election) and the sample. The comparison between the two tables shows that the sample is representative of the population. TABLE 1 shows that the population is 33% Republican, 56% non-Hispanic white, 43% male, and have an average age of 52. And 18% of the population voted by mail. The sample is 31% Republican, 54% non-Hispanic white, 42% male, and have an average age of 50. And 17% of the sample voted by mail. 

Although the sample is representative of the population, I find disparities between the statistics from the state voter file and other external data. The 2019 Census 1-year population estimates show that North Carolina’s total population is 62.5% non-Hispanic white, while TABLE 1 shows that North Carolina’s voting population is 56% non-Hispanic white. Moreover, presidential election voting history shows that North Carolina has voted Republican in nine of the last 10 presidential elections. That implies many North Carolinians who actually vote for the Republican Party are not registered as Republicans. 
For this paper, I disregard these questions due to time constraints, but I intend to explore these issues later. 


# 5. Initial Models
I run a logistic regression with the observational final dataset created by linking various elections datasets from North Carolina. I choose logistic regression because it allows me to work with a binary dependent variable and easily separate the effects of other factors that influence the dependent variable. Whether to VBM is a decision influenced by interactions between various voter characteristics. Therefore, it is necessary to separate the effects of them to measure the sole effect of being Republican on one's probability of VBM. 

I have chosen observational data over survey data because the latter does not have information on each respondent’s residential address and polling location, precluding the calculation and incorporation of the travel distance to a polling location, which is a decisive factor in determining whether to VBM. Voters assigned to a polling location far from their residence have a higher incentive to vote by mail than in person. Therefore, incorporating the travel distance variable in the model is essential. 

Logistic regression assumes (1) linearity in the logit for continuous variables, (2) sample representativeness, (3) absence of multicollinearity, (4) lack of strongly influential outliers, and (5) independence among observations. The first four conditions can be easily met. First, there is no reason to assume non-linearity in the logit for the continuous variables in the model – age, travel distance, and total household income estimate. Second, the individual-level voter file contains information for the entire voter population in North Carolina. Third, no two or more independent variables are strongly correlated to the level where they cannot provide unique information. Fourth, even if several outliers exist, their influence will be minimal due to the large sample size. 

However, the last assumption – independence among observations – can be violated because a person’s decision to vote in person on Election Day is likely to influence the decisions of their family members. For example, a father’s decision to vote in person on Election Day is likely to influence others to join in-person voting by drastically reducing the cost of traveling to the location for the whole family. His family members can simply hop on his car to get to the polling place. The travel can be joyful and even lead to an exciting family outing. 

However, this issue should have been resolved in the sampling process where I randomly selected 10 samples from each precinct. It is very unlikely that two members from one family got selected in the process. 

```{r}
simple_logistic <- glm(VBM ~ Rep, data = data_for_models, family = 'binomial')
summ(simple_logistic)
```
```{r}
stargazer2 <- function(model, odd.ratio = F, ...) {
  if(!("list" %in% class(model))) model <- list(model)
    
  if (odd.ratio) {
    coefOR2 <- lapply(model, function(x) exp(coef(x)))
    seOR2 <- lapply(model, function(x) exp(coef(x)) * summary(x)$coef[, 2])
    p2 <- lapply(model, function(x) summary(x)$coefficients[, 4])
    stargazer(model, coef = coefOR2, se = seOR2, p = p2, ...)
    
  } else {
    stargazer(model, ...)
  }
}
```

**Multiple Logistic Regression without TravelDistance Variable**
```{r}
multiple_logistic_1 <- glm(VBM ~ Rep + Age + White + Male, data = data_for_models, family = 'binomial')
stargazer(multiple_logistic_1, type="html", title="TABLE 2: Logistic Model 1 - Log Odds", align=TRUE, out="model_1.htm")
stargazer(multiple_logistic_1, apply.coef = exp, type="html", title="TABLE 3: Logistic Model 1 - Log Odds (Exponentiated) ", align=TRUE, out="model_1_2.htm")
```

Adding variables the adjusted R-squared scores show that

```{r}
jtools::summ(multiple_logistic_1, exp = TRUE)
```

```{r}
multiple_logistic_2 <- glm(VBM ~ Rep + Age + White + Male + TravelDistance, data = data_for_models, family = 'binomial')
stargazer(multiple_logistic_2, type="html", title="TABLE 4: Logistic Model 2", align=TRUE, out="model_2.htm")
stargazer(multiple_logistic_2, apply.coef = exp, type="html", title="TABLE 5: Logistic Model 2 - Log Odds (Exponentiated) ", align=TRUE, out="model_2_2.htm")
```

**Logistic Regression with Interaction Terms**
```{r}
multiple_logistic_3 <- glm(VBM ~ Rep + Age + White + Male + TravelDistance + Rep*White, data = data_for_models, family = 'binomial')
summary(multiple_logistic_3)
```

# 6. Final Models
```{r}
final_logistic <- glm(VBM ~ Rep + Age + White + Male + TravelDistance + Rep*White, data = data_for_models, family = 'binomial')

stargazer(final_logistic, type="html", title="TABLE 6: Logistic Model 3", align=TRUE, out="model_3.htm")
stargazer(final_logistic, apply.coef = exp, type="html", title="TABLE 7: Logistic Model 3 - Log Odds (Exponentiated) ", align=TRUE, out="model_3_2.htm")
```

```{r}
stargazer(multiple_logistic_1, multiple_logistic_2, final_logistic, title = 'results')
```
```{r}
stargazer(multiple_logistic_1, multiple_logistic_2, final_logistic, type="html", title="TABLE 9: Model Comparison", align=TRUE, out="model_comparison.htm")
```

```{r}
predict(final_logistic, type = "response", newdata = data.frame(Rep = c(1,1), White = c(0,1), Age = c(38,38), Male = c(1,1), TravelDistance=c(1,1)))
```

```{r}
predict(final_logistic, type = "response", newdata = data.frame(Rep = c(0,0), White = c(0,1), Age = c(38,38), Male = c(1,1), TravelDistance=c(1,1)))
```
# Likelihood Ratio Test 1
```{r}
lmtest::lrtest(multiple_logistic_2, multiple_logistic_1)
```

# Likelihood Ratio Test 2
```{r}
lmtest::lrtest(final_logistic, multiple_logistic_2)
```

# Wald Test
```{r}
MANOVA(final_logistic)
```

#  Lagrange multiplier test 
```{r}

```

